---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PlotFTIR

<!-- badges: start -->
[![R-CMD-check](https://github.com/pbulsink/PlotFTIR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pbulsink/PlotFTIR/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/pbulsink/PlotFTIR/graph/badge.svg?token=zrndO2tPwv)](https://codecov.io/gh/pbulsink/PlotFTIR)
<!-- badges: end -->

([Français](#introduction-et-installation))

## Introduction and Installation
The goal of `PlotFTIR` is to easily and quickly kick-start the production of journal-quality FTIR spectral plots in R using ggplot2. The produced plots can be published directly or further modified by ggplot2 functions.

You can install the development version of PlotFTIR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pbulsink/PlotFTIR")
```

## Example

This is a basic example which shows you how to plot a prepared set of FTIR spectra:

```{r basic_plot_en}
library(PlotFTIR)
plot_ftir(sample_spectra)
```

We can also plot spectra in a stacked/offset manner instead of overlaid:
```{r stack_plot_en}
plot_ftir_stacked(biodiesel)
```

## Data Sets

The package contains two datasets to provide example spectra for plotting:
* `biodiesel` is a set of diesels with 0 to 10 % FAMES (biodiesel) content, plus two known and one unknown diesel spectra.
* `sample_spectra` is a set of random FTIR spectra which includes spectra of pure toluene, isopropanol, and heptanes, as well as white printer paper and a polystyrene film.


## Code of Conduct

Please note that the PlotFTIR project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.


([English](#introduction-and-installation))

## Introduction et installation

L'objectif de PlotFTIR est de démarrer facilement et rapidement la production de tracés spectraux FTIR de qualité journal dans R à l'aide de ggplot2. Les tracés produits peuvent être publiés directement ou modifiés davantage par les fonctions ggplot2.

Vous pouvez installer la version de développement de PlotFTIR depuis [GitHub](https://github.com/) avec :

``` r
# install.packages("devtools")
devtools::install_github("pbulsink/PlotFTIR")
```

## Exemple
Ceci est un example de base qui vous montre comment tracer un ensemble prépar de spectres FTIR:

```{r basic_plot_fr}
library(PlotFTIR)
plot_ftir(sample_spectra)
```

Nous pouvons également tracer les spectres de manière empilée/décalée au lieu de les superposer :
```{r stack_plot_fr}
plot_ftir_stacked(biodiesel)
```

## Ensembles de données

Le package contient deux ensembles de données pour fournir des exemples de spectres à tracer :
* `biodiesel` est un ensemble de diesels avec une teneur en FAMES (biodiesel) de 0 à 10 %, plus deux spectres de diesel connus et un inconnu.
* `sample_spectra` est un ensemble de spectres FTIR aléatoires qui comprennent des spectres de toluène pur, d'isopropanol et d'heptanes, ainsi que du papier d'imprimante blanc et un film de polystyrène.


## Code de conduite

Veuillez noter que le projet PlotFTIR est publié avec un [Code de conduite pour le projet](CODE_OF_CONDUCT.md). En contribuant à ce projet, vous acceptez d'en respecter les termes.
