---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(magrittr)
```

# PlotFTIR

<!-- badges: start -->
[![R-CMD-check](https://github.com/pbulsink/PlotFTIR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pbulsink/PlotFTIR/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/pbulsink/PlotFTIR/graph/badge.svg?token=zrndO2tPwv)](https://codecov.io/gh/pbulsink/PlotFTIR)
[![CRAN status](https://www.r-pkg.org/badges/version/PlotFTIR)](https://CRAN.R-project.org/package=PlotFTIR)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

([Français](#introduction-et-installation))

## Introduction and Installation
The goal of `PlotFTIR` is to easily and quickly kick-start the production of journal-quality Fourier Transform Infra-Red (FTIR) spectral plots in R using ggplot2. The produced plots can be published directly or further modified by ggplot2 functions.

You can install the development version of PlotFTIR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("pbulsink/PlotFTIR")
```

## Example Plots

This is a basic example which shows you how to plot a prepared set of FTIR spectra:

```{r basic_plot_en}
library(PlotFTIR)
biodiesel_plot <- plot_ftir(biodiesel)
biodiesel_plot
```

You can also plot spectra in a stacked/offset manner instead of overlaid:
```{r stack_plot_en}
# Generate a plot
plot_ftir_stacked(biodiesel)
```

Note the default plot and legend titles are in english but can be automatically changed to french defaults by supplying the `lang = 'fr'` argument to plot creation functions.

Plots can be manipulated, for example, by zooming in on a range:
```{r biodiesel_zoom_en}
# Zoom to a specified range of 1850 to 1650 cm^-1
zoom_in_on_range(biodiesel_plot, c(1650, 1850))
```

Some FTIR plots have a compressed low-energy portion of the graph which you might wish to zoom in on. You can achieve this by the following:
```{r biodiesel_compress_en}
# compress the data with wavenumbers above 2000 (to the left of 2000 on the
# plot) by a factor of 5
compress_low_energy(biodiesel_plot, cutoff = 2000, compression_ratio = 5)
```

You can also add marker lines (with labels) at specific wavenumbers on the plots, controlling their line or text properties as needed.
```{r biodiesel_labelled_en}
biodiesel_marked <- add_wavenumber_marker(biodiesel_plot,
  wavenumber = 1742,
  text = "C=O Stretch",
  label_aesthetics = list("color" = "red")
)
add_wavenumber_marker(biodiesel_marked,
  wavenumber = 2920,
  text = "C-H Stretch",
  line_aesthetics = list("linetype" = "dashed")
)
```

If the need arises to rename samples listed in the legend, this is possible via `rename_plot_sample_ids()`. At this time, all samples must be listed in the rename vector with the format `"old name" = "new name"`. 
```{r biodiesel_rename_en}
new_names <- c(
  "biodiesel_0" = "0.0% Biodiesel",
  "biodiesel_0_25" = "0.25% Biodiesel",
  "biodiesel_0_50" = "0.50% Biodiesel",
  "biodiesel_1_0" = "1.0% Biodiesel",
  "biodiesel_2_5" = "2.5% Biodiesel",
  "biodiesel_5_0" = "5.0% Biodiesel",
  "biodiesel_7_5" = "7.5% Biodiesel",
  "biodiesel_10_0" = "10.0% Biodiesel",
  "biodiesel_B0_5" = "Commercial B0.5",
  "biodiesel_B5" = "Commercial B5",
  "diesel_unknown" = "Unknown Biodiesel"
)
rename_plot_sample_ids(biodiesel_plot, new_names)
```

A helper function for the renaming is provided (see the documentation for `get_plot_sample_ids()`).

Finally, plot legends are customizable (for basic changes) through a helper function `move_plot_legend()`.

## Data Sets

The package contains two datasets to provide example spectra for plotting:
* `biodiesel` is a set of diesels with 0 to 10 % FAMEs (fatty acid methyl esters; biodiesel) content, plus two known and one unknown diesel spectra.
* `sample_spectra` is a set of random FTIR spectra which includes spectra of pure toluene, isopropanol, and heptanes, as well as white printer paper and a polystyrene film.

An example of the `biodiesel` data set is below:
```{r biodiesel_head_en}
head(biodiesel)
```

## Tidy Plot Production

Note that because most functions return a data type similar to what is provided, tidy-eval is possible (using the base R pipe `|>` or [`magrittr` pipe function](https://magrittr.tidyverse.org/reference/pipe) `%>%`). 
```{r tidy_en, warning=FALSE}
library(magrittr)

new_ids <- c(
  "toluene" = "Toluene", "heptanes" = "C7 Alkane", "isopropanol" = "IPA",
  "paper" = "White Paper", "polystyrene" = "PS Film"
)

sample_spectra |>
  absorbance_to_transmittance() |>
  plot_ftir(plot_title = "Example FTIR Spectra") |>
  zoom_in_on_range(zoom_range = c(3800, 800)) |>
  compress_low_energy(compression_ratio = 4) |>
  add_wavenumber_marker(
    wavenumber = 1495,
    text = "C-C Aromatic",
    line_aesthetics = list("linetype" = "dashed"),
    label_aesthetics = list("color" = "#7e0021")
  ) |>
  add_wavenumber_marker(
    wavenumber = 3340,
    text = "O-H Alcohol",
    line_aesthetics = list("linetype" = "dotted"),
    label_aesthetics = list("color" = "#ff420e")
  ) |>
  rename_plot_sample_ids(sample_ids = new_ids) |>
  move_plot_legend(position = "bottom", direction = "horizontal")
```

## Data Manipulation

FTIR spectral data can be converted between absorbance and transmittance. Only one type of data can exist in a data.frame and be plotted. The functions `absorbance_to_transmittance()` and `transmittance_to_absorbance()` perform these conversions.

```{r convert_datatype_en}
biodiesel_transm <- absorbance_to_transmittance(biodiesel)
head(biodiesel_transm)
```

## Reading Files

PlotFTIR can read .csv and .asp file types. The .csv file should contain only one spectra, with columns for `wavenumber` and `absorbance` or `transmittance.` The .asp files should be according to the file specifications (not modified by the user).

## Documentation

More information on all functions is available online at [https://plotftir.pages.dev/](https://plotftir.pages.dev/).

## Code of Conduct

Please note that the PlotFTIR project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

## Citing This Package

Please cite this package in any journal articles containing images produced by way of the package. If installed from GitHub or CRAN the date field will be properly filled with the publishing year.

```{r cite-en, warning=FALSE}
citation("PlotFTIR")
```

([English](#introduction-and-installation))

## Introduction et installation

Le but de `PlotFTIR` est de lancer facilement et rapidement la production des tracés de spectres de spectroscopie infrarouge à transformée de Fourier (IRTF) de qualité de revues scientifiques dans le system R en utilisant ggplot2. Les tracés produits peuvent être publiés directement ou modifiés par les fonctions ggplot2.

Vous pouvez installer la version de développement de `PlotFTIR` depuis [GitHub](https://github.com/) avec:

``` r
# install.packages("devtools")
devtools::install_github("pbulsink/PlotFTIR")
```

## Exemples des tracés
Ceci est un example de base qui vous montre comment tracer un ensemble de spectres IRTF dejà preparé:

```{r basic_plot_fr}
library(PlotFTIR)
plot_ftir(sample_spectra, lang = "fr")
```

Vous pouvez également tracer les spectres de manière empilée/décalée au lieu de les superposer :
```{r stack_plot_fr}
plot_ftir_stacked(biodiesel, plot_title = "Spectre IRTF empilée", lang = "fr")
```

Notez que les titres par défaut de tracé et du légende sont en anglais, mais qu'ils peuvent être automatiquement modifiés en français en fournissant l'argument `lang = 'fr'` aux fonctions de création de tracés.

Les tracés peuvent être manipulés, par exemple, en zoomant sur une plage :
```{r biodiesel_zoom_fr}
# Générer un tracé
biodiesel_trace <- plot_ftir(biodiesel, lang = "fr")
# Zoom sur une plage spécifiée de 1850 à 1650 cm^-1
zoom_in_on_range(biodiesel_trace, c(1650, 1850))
```

Certains tracés IRTF ont une partie compressée du graphique à faible énergie qui peuvent etre agrandie de la manière suivante :
```{r biodiesel_compress_fr}
# compresser les données avec des nombres d'onde supérieurs à 2000 (à gauche de
# 2000 sur le tracé) d'un facteur 5
compress_low_energy(biodiesel_trace, cutoff = 2000, compression_ratio = 5)
```

Vous pouvez également ajouter des lignes de marqueur (avec des étiquettes) à des numéros d'onde spécifiques sur les tracés, en contrôlant leurs propriétés de ligne ou de texte selon vos besoins.
```{r biodiesel_labelled_fr}
biodiesel_marked <- add_wavenumber_marker(biodiesel_trace,
  wavenumber = 1742,
  text = "C=O étirement",
  label_aesthetics = list("color" = "red")
)
add_wavenumber_marker(biodiesel_trace,
  wavenumber = 2920,
  text = "C-H étirement",
  line_aesthetics = list("linetype" = "dashed")
)
```

S'il est nécessaire de renommer les échantillons répertoriés dans la légende, cela est possible via `rename_plot_sample_ids()`. À ce stade, tous les échantillons doivent être répertoriés dans le vecteur de renommage au format `"ancien nom" = "nouveau nom"`.

```{r biodiesel_rename_fr}
nouveau_noms <- c(
  "biodiesel_0" = "0,0% Biodiesel",
  "biodiesel_0_25" = "0,25% Biodiesel",
  "biodiesel_0_50" = "0,50% Biodiesel",
  "biodiesel_1_0" = "1,0% Biodiesel",
  "biodiesel_2_5" = "2,5% Biodiesel",
  "biodiesel_5_0" = "5,0% Biodiesel",
  "biodiesel_7_5" = "7,5% Biodiesel",
  "biodiesel_10_0" = "10,0% Biodiesel",
  "biodiesel_B0_5" = "B0,5 Commercial",
  "biodiesel_B5" = "B5 Commercial",
  "diesel_unknown" = "Biodiesel Inconnu"
)
rename_plot_sample_ids(biodiesel_trace, nouveau_noms)
```

Une fonction d'assistance pour le changement de nom est fournie (voir la documentation pour `get_plot_sample_ids()`).

Enfin, les légendes des tracés sont personnalisables (pour les modifications de base) via une fonction d'assistance `move_plot_legend()`.

## Production de tracés "tidy" 

Notez que comme la plupart des fonctions renvoient un type de données similaire à celui qui est fourni, l'évaluation des données est possible (en utilisant [la fonction `magrittr` tuyau](https://magrittr.tidyverse.org/reference/pipe) `%>%`).

```{r tidy_fr, warning=FALSE}
library(magrittr)

nouveaux_ids <- c(
  "toluene" = "Toluène", "heptanes" = "C7 alcane", "isopropanol" = "alcool isopropylique",
  "paper" = "papier blanc", "polystyrene" = "film de polystyrène"
)

sample_spectra |>
  absorbance_to_transmittance() |>
  plot_ftir(plot_title = "Exemple de spectres IRTF", lang = "fr") |>
  zoom_in_on_range(zoom_range = c(3800, 800)) |>
  compress_low_energy(compression_ratio = 4) |>
  add_wavenumber_marker(
    wavenumber = 1495,
    text = "C-C aromatique",
    line_aesthetics = list("linetype" = "dashed"),
    label_aesthetics = list("color" = "#7e0021")
  ) |>
  add_wavenumber_marker(
    wavenumber = 3340,
    text = "O-H alcool",
    line_aesthetics = list("linetype" = "dotted"),
    label_aesthetics = list("color" = "#ff420e")
  ) |>
  rename_plot_sample_ids(sample_ids = nouveaux_ids) |>
  move_plot_legend(position = "bottom", direction = "horizontal")
```

## Ensembles des données

Le package contient deux ensembles de données pour fournir des exemples de spectres à tracer:
* `biodiesel` est un ensemble de diesels avec une teneur en esters méthyliques d'acides gras (EMAGs) (ou biodiesel) de 0 à 10 %, plus deux spectres de diesel connus et un inconnu.
* `sample_spectra` est un ensemble de spectres IRTF aléatoires qui comprennent des spectres de toluène pur, d'isopropanol et d'heptanes, ainsi que du papier d'imprimante blanc et un film de polystyrène.

Un exemple de l'ensemble de données `biodiesel` est ci-dessous:
```{r biodiesel_head_fr}
head(biodiesel)
```

## Manipulation de données

Les données spectrales IRTF peuvent être converties entre l'absorbance et la transmission. Un seul type de données peut exister dans un data.frame et être tracé. Les fonctions `absorbance_to_transmittance()` et `transmittance_to_absorbance()` effectuent ces conversions.

```{r convert_datatype_fr}
biodiesel_transm <- absorbance_to_transmittance(biodiesel)
head(biodiesel_transm)
```

## Lecture des fichiers

PlotFTIR peut lire les fichiers de type .csv et .asp. Le fichier .csv ne doit contenir qu'un seul spectre, avec des colonnes pour le `wavenumber` et `absorbance` ou `transmittance`. Les fichiers .asp doivent être conformes aux spécifications du fichier (non modifiées par l'utilisateur).

## Documentation

Plus d'informations sur toutes les fonctions sont disponibles en ligne sur [https://plotftir.pages.dev/](https://plotftir.pages.dev/).

## Code de conduite

Veuillez noter que le projet PlotFTIR est publié avec un [Code de conduite pour le projet](CODE_OF_CONDUCT.md). En contribuant à ce projet, vous acceptez de respecter ces conditions.


## Citer ce paquet

Veuillez citer ce paquet dans tout article de journal contenant des images produites à l'aide de ce paquet. Si le paquet est installé à partir de GitHub ou de CRAN, le texte de la date sera correctement rempli avec l'année de publication.

```{r cite-fr, warning=FALSE}
citation("PlotFTIR")
```
